---
title: "Analytic Sample"
format: html
editor: visual
author: Jill Kumasaka
date: today
---

## load packages

```{r}
#| echo: false

library(haven)
library(dplyr)
library(here)
library(mosaic)
library(psych)
library(questionr)
library(rstatix)
here()
```


## 2011-2012 data merge

```{r}
#generate temporary files for 2011-2012
temp_AUQ_G <- read_xpt(here("Data/AUQ_G.xpt"))
temp_AUX_G <- read_xpt(here("Data/AUX_G.xpt"))
temp_DEMO_G <- read_xpt(here("Data/DEMO_G.xpt"))
temp_SMQ_G <- read_xpt(here("Data/SMQ_G.xpt"))
temp_VOCWB_G <- read_xpt(here("Data/VOCWB_G.xpt"))

#merge files for 2011-2012
merged_1112 <- left_join(temp_DEMO_G, temp_AUQ_G, by= "SEQN")
merged_1112 <- left_join(merged_1112, temp_SMQ_G, by= "SEQN")
merged_1112 <- left_join(merged_1112, temp_AUX_G, by= "SEQN")
merged_1112 <- left_join(merged_1112, temp_VOCWB_G, by= "SEQN")

#check that number of rows from original matches number of rows in merged
nrow(temp_DEMO_G) == nrow(merged_1112) #true - successful merge
```

## 2015-2016 data merge

```{r}
#generate temporary files for 2015-2016
temp_AUQ_I <- read_xpt(here("Data/AUQ_I.xpt"))
temp_AUX_I <- read_xpt(here("Data/AUX_I.xpt"))
temp_DEMO_I <- read_xpt(here("Data/DEMO_I.xpt"))
temp_SMQ_I <- read_xpt(here("Data/SMQ_I.xpt"))
temp_VOCWB_I <- read_xpt(here("Data/VOCWB_I.xpt"))

#merge files for 2015-2016
merged_1516 <- left_join(temp_DEMO_I, temp_AUQ_I, by= "SEQN")
merged_1516 <- left_join(merged_1516, temp_SMQ_I, by= "SEQN")
merged_1516 <- left_join(merged_1516, temp_AUX_I, by= "SEQN")
merged_1516 <- left_join(merged_1516, temp_VOCWB_I, by= "SEQN")

#check that number of rows from original matches number of rows in merged
nrow(temp_DEMO_I) == nrow(merged_1516) #true - successful merge

```

## append data

```{r}
#add the observations from each dataset to each other
appended_1112_1516<-bind_rows(merged_1112,
                              merged_1516)
```

## add survey weights

```{r}
#use VOC bc least common denominator 
appended_1112_1516 <- appended_1112_1516 %>% 
  mutate(WTSVOC4Y = WTSVOC2Y/2)
```

## save appended data

```{r}
save(appended_1112_1516, file = "appended_dataset.rda") #19,727 observations 
```

## create age variable 20-69 years

```{r}
appended_edited <- appended_1112_1516 %>%
  mutate(age_instudy = case_match(
    RIDAGEYR, c(20:69) ~ 1, 
    RIDAGEYR < 20 ~ 0, 
    RIDAGEYR > 69 ~ 0,
    .default = 0
  ))

freq(appended_edited$age_instudy) #9,444 observations in age range 

```

## data exploration - occupational noise exposure

```{r}
#create occupational noise exposure variable bc 2011-12 and 2015-16 is different variable name
appended_edited <- appended_edited %>%
  mutate(
    occ_noise = coalesce(AUQ330, AUQ331))
table(appended_edited$occ_noise)

#recode so yes=1 and no=0
appended_edited <- appended_edited %>%
  mutate(
    occ_noise = case_when(
      occ_noise == 1 ~ 1,
      occ_noise == 2 ~ 0
    )
  )
appended_edited %>% freq_table(occ_noise)
```


## create variables for missing data

```{r}
#missing data on outcomes
appended_edited <- appended_edited %>%
  filter(age_instudy == 1) %>%
  mutate(miss_out = case_when(is.na(AUXU3KR) |
                                is.na(AUXU4KR) |
                                is.na(AUXU6KR) |
                                is.na(AUXU8KR) |
                                is.na(AUXU3KL) |
                                is.na(AUXU4KL) |
                                is.na(AUXU3KL) |
                                is.na(AUXU3KL) ~ 1,
                             .default = 0))

table(appended_edited$miss_out)

#missing data on exposure 
#including only participants with all VOCs measured - excluding toluene
#if we want to include participants who have at least one VOC, use "&" instead of "|" logic
appended_edited <- appended_edited %>%
  filter(age_instudy == 1) %>%
  mutate(miss_exp = case_when(is.na(LBXVBZ) |
                                is.na(LBXVEB) |
                                is.na(LBXVOX) |
                                is.na(LBXVXY) | 
                                is.na(occ_noise) ~ 1,
                             TRUE ~ 0))

table(appended_edited$miss_exp)

#missing data on EMM
appended_edited <- appended_edited %>%
  filter(age_instudy == 1) %>%
  mutate(miss_EMM = case_when(is.na(RIDRETH3) ~ 1,
                             TRUE ~ 0))

table(appended_edited$miss_EMM)

#missing data on confounders
appended_edited <- appended_edited %>%
  filter(age_instudy == 1) %>%
  mutate(miss_confound = case_when(is.na(RIDAGEYR) |
                                is.na(DMDEDUC2) |
                                is.na(INDFMPIR) |
                                is.na(RIAGENDR) |
                                is.na(SMQ020) ~ 1,
                             TRUE ~ 0))
table(appended_edited$miss_confound)

```

## subpopulation of only complete cases in age range

```{r}
#create subpopulation variable of only age included and complete cases 
appended_edited <- appended_edited %>%
  mutate(subpop1 = case_when(age_instudy == 1 & 
                               miss_out == 0 & 
                               miss_exp == 0 & 
                               miss_EMM == 0 & 
                               miss_confound == 0 ~ 1, 
                             .default = 0))
table(appended_edited$subpop1)

#create dataset of only age included and complete cases 
appended_complete <- appended_edited %>%
  filter(subpop1 == 1)
```


## data exploration & create binary exposure - benzene

```{r}
#benzene
summary(appended_complete$LBXVBZ)
#histogram
hist(
  appended_complete$LBXVBZ,
  breaks = 80,
  main = "Histogram of Continous Blood Benzene",
  xlab = "LBXVBZ",
  col = "lightblue",
  border = "black"
)
#create binary lod variable (LOD=0.024 - value is 0.017)
appended_complete <- appended_complete %>%
  mutate(benzene_nolod = case_when(
    LBXVBZ > 0.017 ~ 1, 
    LBXVBZ <= 0.017 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$benzene_nolod)

#bar graph
ggplot(appended_complete, aes(x = benzene_nolod)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary Benzene Using LOD",
    x = "0=below LOD; 1=above LOD",
    y = "Count"
  ) +
  theme_minimal()

#subpopulation with only values abov LOD
subpop_benzene_lod <- appended_complete %>%
  filter(benzene_nolod == 1)


#create quartiles with values above LOD
subpop_benzene_lod <- subpop_benzene_lod %>%
  mutate(
    benzene_quart = ntile(LBXVBZ, 4)
  )

#label quartiles
subpop_benzene_lod <- subpop_benzene_lod %>%
  mutate(
    benzene_quart = factor(
      benzene_quart,
      levels = 1:4,
      labels = c("Q1", "Q2", "Q3", "Q4")
    )
  )

#check quartile values
subpop_benzene_lod %>%
  group_by(benzene_quart) %>%
  summarise(
    n = n(),
    min = min(LBXVBZ, na.rm = TRUE),
    max = max(LBXVBZ, na.rm = TRUE)
  )
    #Q3 min = 0.08 threshold

#create binary benzene variable with T3 cutoff
appended_complete <- appended_complete %>%
  mutate(benzene_binary = case_when(
    LBXVBZ >= 0.08 ~ 1, 
    LBXVBZ < 0.08 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$benzene_binary)

#bar graph
ggplot(appended_complete, aes(x = benzene_binary)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary Benzene Using Quartiles",
    x = "0=below Q3 threshold; 1=above Q3 threshold",
    y = "Count"
  ) +
  theme_minimal()

#cross-tabulate with noise exposure
table(appended_complete$benzene_binary, appended_complete$occ_noise)

```


## data exploration & create binary exposure - ethylbenzene

```{r}
#ethylbenzene
summary(appended_complete$LBXVEB)
#histogram
hist(
  appended_complete$LBXVEB,
  breaks = 80,
  main = "Histogram of Continous Blood Ethylbenzene",
  xlab = "LBXVBZ",
  col = "lightblue",
  border = "black"
)
#create binary lod variable (LOD=0.024 - value is 0.017)
appended_complete <- appended_complete %>%
  mutate(ethylbenzene_nolod = case_when(
    LBXVEB > 0.017 ~ 1, 
    LBXVEB <= 0.017 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$ethylbenzene_nolod)

#bar graph
ggplot(appended_complete, aes(x = ethylbenzene_nolod)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary Ethylbenzene Using LOD",
    x = "0=below LOD; 1=above LOD",
    y = "Count"
  ) +
  theme_minimal()

#subpopulation with only values above LOD
subpop_ethylbenzene_lod <- appended_complete %>%
  filter(ethylbenzene_nolod == 1)


#create quartiles with values above LOD
subpop_ethylbenzene_lod <- subpop_ethylbenzene_lod %>%
  mutate(
    ethylbenzene_quart = ntile(LBXVEB, 4)
  )

#label quartiles
subpop_ethylbenzene_lod <- subpop_ethylbenzene_lod %>%
  mutate(
    ethylbenzene_quart = factor(
      ethylbenzene_quart,
      levels = 1:4,
      labels = c("Q1", "Q2", "Q3", "Q4")
    )
  )

#check quartile values
subpop_ethylbenzene_lod %>%
  group_by(ethylbenzene_quart) %>%
  summarise(
    n = n(),
    min = min(LBXVEB, na.rm = TRUE),
    max = max(LBXVEB, na.rm = TRUE)
  )
    #Q3 min = 0.058 threshold

#create binary ethylbenzene variable with Q3 cutoff
appended_complete <- appended_complete %>%
  mutate(ethylbenzene_binary = case_when(
    LBXVEB >= 0.058 ~ 1, 
    LBXVEB < 0.058 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$ethylbenzene_binary)

#bar graph
ggplot(appended_complete, aes(x = ethylbenzene_binary)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary Ethylbenzene Using Quartiles",
    x = "0=below Q3 threshold; 1=above Q3 threshold",
    y = "Count"
  ) +
  theme_minimal()

#cross-tabulate with noise exposure
table(appended_complete$ethylbenzene_binary, appended_complete$occ_noise)

```

## data exploration & create binary exposure - o-xylene
```{r}
#o-xylene
summary(appended_edited$LBXVOX)
#histogram
hist(
  appended_edited$LBXVOX,
  breaks = 80,
  main = "Histogram of O-xylene",
  xlab = "LBXVOX",
  col = "lightblue",
  border = "black"
)

#create binary lod variable (LOD=0.024 - value is 0.017)
appended_complete <- appended_complete %>%
  mutate(oxylene_nolod = case_when(
    LBXVOX > 0.017 ~ 1, 
    LBXVOX <= 0.017 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$oxylene_nolod)

#bar graph
ggplot(appended_complete, aes(x = oxylene_nolod)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary O-xylene Using LOD",
    x = "0=below LOD; 1=above LOD",
    y = "Count"
  ) +
  theme_minimal()

#subpopulation with only values above LOD
subpop_oxylene_lod <- appended_complete %>%
  filter(oxylene_nolod == 1)

#create quartiles with values above LOD
subpop_oxylene_lod <- subpop_oxylene_lod %>%
  mutate(
    oxylene_quart = ntile(LBXVOX, 4)
  )

#label quartiles
subpop_oxylene_lod <- subpop_oxylene_lod %>%
  mutate(
    oxylene_quart = factor(
      oxylene_quart,
      levels = 1:4,
      labels = c("Q1", "Q2", "Q3", "Q4")
    )
  )

#check quartile values
subpop_oxylene_lod %>%
  group_by(oxylene_quart) %>%
  summarise(
    n = n(),
    min = min(LBXVOX, na.rm = TRUE),
    max = max(LBXVOX, na.rm = TRUE)
  )
    #Q3 min = 0.042 threshold

#create binary o-xylene variable with Q3 cutoff
appended_complete <- appended_complete %>%
  mutate(oxylene_binary = case_when(
    LBXVOX >= 0.042 ~ 1, 
    LBXVOX < 0.042 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$oxylene_binary)

#bar graph
ggplot(appended_complete, aes(x = oxylene_binary)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary O-xylene Using Quartiles",
    x = "0=below Q3 threshold; 1=above Q3 threshold",
    y = "Count"
  ) +
  theme_minimal()

#cross-tabulate with noise exposure
table(appended_complete$oxylene_binary, appended_complete$occ_noise) #221 with both exposures

```


## data exploration & create binary exposure - m-/p-xylene ***

```{r}
#m-/p-xylene
summary(appended_edited$LBXVXY)
#histogram
hist(
  appended_edited$LBXVXY,
  breaks = 1000,
  main = "Histogram of m-/p-xylene",
  xlab = "LBXVXY",
  col = "lightblue",
  border = "black"
)

#create binary lod variable (LOD=0.034 - value is 0.024)
appended_complete <- appended_complete %>%
  mutate(mpxylene_nolod = case_when(
    LBXVXY > 0.024 ~ 1, 
    LBXVXY <= 0.024 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$mpxylene_nolod)

#bar graph
ggplot(appended_complete, aes(x = mpxylene_nolod)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary m/p-xylene Using LOD",
    x = "0=below LOD; 1=above LOD",
    y = "Count"
  ) +
  theme_minimal()

#subpopulation with only values above LOD
subpop_mpxylene_lod <- appended_complete %>%
  filter(mpxylene_nolod == 1)

#create quartiles with values above LOD
subpop_mpxylene_lod <- subpop_mpxylene_lod %>%
  mutate(
    mpxylene_quart = ntile(LBXVXY, 4)
  )

#label quartiles
subpop_mpxylene_lod <- subpop_mpxylene_lod %>%
  mutate(
    mpxylene_quart = factor(
      mpxylene_quart,
      levels = 1:4,
      labels = c("Q1", "Q2", "Q3", "Q4")
    )
  )

#check quartile values
subpop_mpxylene_lod %>%
  group_by(mpxylene_quart) %>%
  summarise(
    n = n(),
    min = min(LBXVXY, na.rm = TRUE),
    max = max(LBXVXY, na.rm = TRUE)
  )
    #Q3 min = 0.075 threshold

#create binary m-/p-xylene variable with Q3 cutoff
appended_complete <- appended_complete %>%
  mutate(oxylene_binary = case_when(
    LBXVOX >= 0.075 ~ 1, 
    LBXVOX < 0.075 ~ 0, 
    TRUE ~ 0
  ))
table(appended_complete$oxylene_binary)

#bar graph
ggplot(appended_complete, aes(x = oxylene_binary)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Bar Chart of Binary O-xylene Using Quartiles",
    x = "0=below Q3 threshold; 1=above Q3 threshold",
    y = "Count"
  ) +
  theme_minimal()

#cross-tabulate with noise exposure
table(appended_complete$oxylene_binary, appended_complete$occ_noise) #70 with both exposures - consider tertiles

```


## create outcome audiometric testing binary variable 

```{r}
#make variables for average of right and left ears 
appended_complete <- appended_complete %>%
  mutate(
    PTA_right = rowMeans(select(., AUXU3KR, AUXU4KR, AUXU6KR, AUXU8KR), na.rm = FALSE),
    PTA_left  = rowMeans(select(., AUXU3KL, AUXU4KL, AUXU6KL, AUXU8KL), na.rm = FALSE),
    PTA_better = pmin(PTA_right, PTA_left, na.rm = FALSE))

appended_complete <- appended_complete %>%
  mutate(hearing_loss = if_else(PTA_better > 20, 1, 0))

table(appended_complete$hearing_loss)
```

## recode gender

```{r}
#male=1, female=0
appended_complete <- appended_complete %>%
  mutate(
    RIAGENDR = case_when(
      RIAGENDR == 1 ~ 1,
      RIAGENDR == 2 ~ 0
    )
  )
appended_complete %>% freq_table(RIAGENDR)
```


## recode smoking status

```{r}
#smoker=1, non-smoker=0
appended_complete <- appended_complete %>%
  mutate(
    SMQ020 = case_when(
      SMQ020 == 1 ~ 1,
      SMQ020 == 2 ~ 0
    )
  )
appended_complete %>% freq_table(SMQ020)
```

## save complete dataset
```{r}
#i chose a weird name for the dataset so renaming to complete_dataset for datset with all complete cases 
complete_dataset <- appended_complete
save(complete_dataset, file = "complete_dataset.rda") 
```